---
title: "Simulating Field Trial Data and Visualising Spatial Patterns"
subtitle: "Usage guide for `simulate_spatial_data.R`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5,
  dpi     = 150
)
```

## Overview

`simulate_spatial_data.R` provides three functions for generating and visualising
synthetic multi-environment field trial data. The simulated data is intended
as a test input for the spatial correction pipeline in `SpatialCorrectionFlexible.R`.

**Experimental design simulated:**

- Unreplicated multi-bench trial: each genotype appears once per bench
- Each bench is laid out on a regular row × column grid
- The observed phenotype is the sum of a true genotype effect, a configurable
  spatial environmental gradient, and random plot-level noise

**Functions provided:**

| Function | Purpose |
|---|---|
| `simulate_field_trial()` | Generate a synthetic dataset |
| `plot_bench_heatmap()` | Heatmap for a single bench |
| `plot_all_bench_heatmaps()` | Composite heatmap for all benches, saved as PNG |

---

## Setup

Source the script to load all three functions into your session.

```{r source}
source("../scripts/simulate_spatial_data.R")
```

---

## Function reference

### `simulate_field_trial()`

Generates a complete multi-bench dataset. Returns a list with three elements:

- `$data`: the plot-level data frame (`Bench`, `Genotype`, `Row`, `Col`, `True_Spatial`, `<pheno_name>`)
- `$true_effects`: the underlying genotype effects used to generate the data (`Genotype`, `True_<pheno_name>`)
- `$pheno_name`: the name of the phenotype column (useful for passing to plot functions)

#### Parameters

The number of genotypes is always derived as `n_rows × n_cols`, giving exactly
one replicate per genotype per bench. There is no `n_geno` argument; changing
bench dimensions automatically sets the number of genotypes.

| Parameter | Default | Description |
|---|---|---|
| `n_bench` | `4` | Number of benches/fields. |
| `n_rows` | `10` | Rows per bench. |
| `n_cols` | `30` | Columns per bench. `n_rows × n_cols` genotypes are simulated. |
| `mean_pheno` | `55` | Grand mean of the phenotype (centre of the genotype distribution). |
| `sd_geno` | `12` | SD of true genotype effects: larger values produce more spread between genotypes. |
| `spatial_type` | `5` | Spatial pattern type (see table below). |
| `spatial_intensity` | `30` | Peak magnitude of the spatial gradient in phenotype units, interpreted relative to `sd_geno`: **0** = no effect; **< sd_geno** = subtle (spatial < genetic variation, BLUPs easily recovered); **~ sd_geno** = moderate (realistic for most trials); **> sd_geno** = strong (spatial dominates, correction critical); **>> sd_geno** = extreme (observed values likely exceed `[pheno_min, pheno_max]`). |
| `spatial_scale` | `1` | Steepness/bandwidth multiplier. Types 2, 3, 4: `1` = full gradient from edge to edge; `< 1` = shallower; `> 1` = peaks before the far edge then descends. Type 5: `> 1` = broader Gaussian patch; `< 1` = sharper spotlight. Type 6: `> 1` = sharper edge band; `< 1` = broader decay into interior. Type 7: `> 1` = wider hotspot; `< 1` = more compact. |
| `spatial_intensity_per_bench` | `NULL` | Numeric vector of length `n_bench`. When supplied, overrides `spatial_intensity` with a different peak magnitude for each bench. Useful for testing heterogeneous spatial severity (e.g., `c(15, 25, 40, 60)`). |
| `spatial_type_per_bench` | `NULL` | Integer vector of length `n_bench`. When supplied, overrides `spatial_type` with a different pattern type for each bench. |
| `sd_error` | `4` | Plot-level random noise SD (measurement and microenvironment error). |
| `pheno_min` | `0` | Lower bound: true genotype effects are clipped to `[pheno_min, pheno_max]`. |
| `pheno_max` | `100` | Upper bound for genotype effect clipping. |
| `pheno_name` | `"BNI"` | Name given to the phenotype column in the output. |
| `save_csv` | `TRUE` | If `TRUE`, writes `<pheno_name>_simulation.csv` and `<pheno_name>_true_effects.csv` to `output_dir`. |
| `output_dir` | `"data"` | Directory for CSV output; created automatically if it does not exist. |
| `seed` | `42` | Random seed for reproducibility. |

#### Spatial pattern types (`spatial_type`)

| Value | Pattern | Centred | Description |
|---|---|---|---|
| `1` | No spatial effect | - | 0 everywhere |
| `2` | Row gradient | Yes | Starts at 0 at row 1, rises toward the last row (half-sine); uniform across columns |
| `3` | Column gradient | Yes | Starts at 0 at col 1, rises toward the last column (half-sine); uniform across rows |
| `4` | Row + column gradient | Yes | Additive combination of types 2 and 3; `spatial_intensity` split equally between directions |
| `5` | Central patch | Yes | Gaussian spotlight centred on the bench; sigma scaled to bench dimensions (`σ = n/4 × spatial_scale`) |
| `6` | Edge effects | Yes | Elevated at bench borders with diagonal asymmetry (top-left high, bottom-right low); mimics glasshouse edge drying, irrigation run-off, and asymmetric light/wind exposure |
| `7` | Localised hotspot | Yes | Gaussian bell at an off-centre position (~35% along rows, ~65% along columns); mimics a point-source disturbance such as a roof leak or faulty irrigation emitter, affecting ~20% of pots at `spatial_scale = 1` |

**Centring:** All types except type 1 are centred to mean zero per bench (`raw - mean(raw)`). This ensures the spatial component represents deviations from the bench mean rather than an absolute level shift, which is the correct parameterisation for joint multi-bench models where the spatial smooth is zero-mean constrained.

The data-generating model for each plot is:

$$y_{ij} = g_i + s_j + \varepsilon_{ij}$$

where $g_i \sim N(\mu_g, \sigma_g^2)$ clipped to [`pheno_min`, `pheno_max`]
is the true genotype effect (`mean_pheno`, `sd_geno`),
$s_j$ is the spatial contribution at plot $j$ (determined by `spatial_type` and `spatial_intensity`, centred to mean zero),
and $\varepsilon_{ij} \sim N(0, \sigma_e^2)$ is random plot-level noise (`sd_error`).
The true genotype effects $g_i$ are stored in `$true_effects`, and the per-plot spatial contributions are stored in the `True_Spatial` column of `$data`. Both are used after spatial correction to validate BLUE/BLUP recovery via Pearson correlation and RMSE.
When `save_csv = TRUE` (default), `<pheno_name>_simulation.csv` and
`<pheno_name>_true_effects.csv` are written automatically to `output_dir`.

---

### `plot_bench_heatmap()`

Plots the field layout for a **single bench** as a coloured tile grid.
Returns a `ggplot` object that can be further customised with `ggplot2` layers
or composed into multi-panel figures with `patchwork`.

#### Parameters

| Parameter | Default | Description |
|---|---|---|
| `bench_dfr` | none | Data frame containing the bench data. Pass the full dataset when using `bench`, or a pre-subsetted single-bench frame when `bench = NULL`. |
| `pheno` | none | Name of the response column to use as the fill colour. |
| `rn` | `"Row"` | Name of the row position column. |
| `cn` | `"Col"` | Name of the column position column. |
| `gt` | `"Genotype"` | Name of the genotype column (only used when `show_labels = TRUE`). |
| `bench_col` | `"Bench"` | Name of the bench identifier column; used only when `bench` is specified. |
| `bench` | `NULL` | Bench identifier to select from `bench_dfr`. When set, the function subsets the data internally (no pre-filtering needed). |
| `bench_label` | `""` | Title displayed above the panel. Defaults to the `bench` value when `bench` is specified. |
| `show_labels` | `FALSE` | If `TRUE`, genotype names are printed inside each tile. |
| `shared_limits` | `NULL` | `c(min, max)` for the fill colour scale. Pass the global range across all benches to make scales comparable. `NULL` = auto-scale to this bench only. |
| `label_size` | `2.5` | Font size of genotype labels (in `ggplot2` units). |

**Axis orientation:** column 1 is at the left, last column at the right
(x-axis, ascending). Row 1 is at the top, last row at the bottom
(y-axis reversed to match field convention). This matches the physical
layout of a bench when viewed from above.

---

### `plot_all_bench_heatmaps()`

Loops over all benches, assembles panels with `patchwork`, and saves the
composite figure as a PNG file.

#### Parameters

| Parameter | Default | Description |
|---|---|---|
| `dfr` | none | Full data frame (all benches). |
| `pheno` | none | Name of the response column. |
| `rn` | `"Row"` | Row column name. |
| `cn` | `"Col"` | Column column name. |
| `gt` | `"Genotype"` | Genotype column name. |
| `bench_col` | `"Bench"` | Column identifying the bench. |
| `show_labels` | `FALSE` | Show genotype names on tiles. |
| `shared_scale` | `TRUE` | Use a common colour scale across all benches. |
| `output_png` | `"heatmaps.png"` | Output file path. |
| `ncol_panels` | `2` | Number of bench panels per row in the composite. |
| `panel_width` | `7` | Width per panel in inches. |
| `panel_height` | `4` | Height per panel in inches. |
| `label_size` | `2.5` | Font size for genotype labels. |

---

## Worked example

### 1. No spatial effect (type 1)

A baseline simulation with no environmental gradient. All observed variation
reflects genotype differences and random plot noise only.

```{r sim-none}
sim_none <- simulate_field_trial(
  spatial_type      = 1,
  spatial_intensity = 0,
  pheno_name        = "BNI",
  output_dir        = "data",
  seed              = 42
)

head(sim_none$data)
```

```{r sim-none-plot, fig.height=4}
plot_all_bench_heatmaps(
  dfr          = sim_none$data,
  pheno        = "BNI",
  bench_col    = "Bench",
  show_labels  = FALSE,
  shared_scale = TRUE,
  output_png   = "data/heatmap_type1_none.png"
)
```

With no spatial effect the colour pattern across each bench is essentially
random, so no structure is visible.

---

### 2. Column gradient (type 3)

The column gradient starts at 0 at column 1 and rises toward the last column
(half-sine curve). This mimics a light or temperature gradient running the
length of a glasshouse bench.

```{r sim-cols}
sim_cols <- simulate_field_trial(
  spatial_type      = 3,
  spatial_intensity = 30,
  pheno_name        = "BNI",
  output_dir        = "data",
  seed              = 42
)

head(sim_cols$data)
cat("Observations:", nrow(sim_cols$data), "\n")
cat("Genotypes:   ", length(unique(sim_cols$data$Genotype)), "\n")
cat("Benches:     ", length(unique(sim_cols$data$Bench)), "\n")
```

```{r sim-cols-plot, fig.height=4}
plot_all_bench_heatmaps(
  dfr          = sim_cols$data,
  pheno        = "BNI",
  bench_col    = "Bench",
  show_labels  = FALSE,
  shared_scale = TRUE,
  output_png   = "data/heatmap_type3_cols.png"
)
```

Vertical bands of colour are visible, with the lowest values at column 1 and
the highest at the last column, forming a clear monotonic column-wise gradient.

---

### 3. Multi-bench heatmaps

`plot_all_bench_heatmaps()` loops over all benches, assembles the panels with
`patchwork`, and saves the result as a PNG. The key layout parameters are
`ncol_panels` (bench panels per row), `panel_width`, and `panel_height`.

The default `ncol_panels = 2` arranges panels two-per-row, which is compact
for reports:

```{r multi-bench-2col, fig.height=9, fig.width=14}
p_all <- plot_all_bench_heatmaps(
  dfr          = sim_cols$data,
  pheno        = "BNI",
  bench_col    = "Bench",
  show_labels  = FALSE,
  shared_scale = TRUE,
  ncol_panels  = 2,
  panel_width  = 7,
  panel_height = 4,
  output_png   = "data/heatmap_type3_cols_2col.png"
)
p_all
```

Setting `ncol_panels = 1` arranges benches in a single column so each fills
the full width, which is useful when benches are wide (many columns) and side-by-side
panels would be too narrow to read:

```{r multi-bench-1col, fig.height=16, fig.width=9}
p_all1 <- plot_all_bench_heatmaps(
  dfr          = sim_cols$data,
  pheno        = "BNI",
  bench_col    = "Bench",
  show_labels  = FALSE,
  shared_scale = TRUE,
  ncol_panels  = 1,
  panel_width  = 9,
  panel_height = 4,
  output_png   = "data/heatmap_type3_cols_1col.png"
)
p_all1
```

---

### 4. Plotting a single bench

`plot_bench_heatmap()` returns a `ggplot` object for a single bench,
allowing further customisation or composition with `patchwork`. Pass the full
dataset and specify `bench` to select which bench to plot (no pre-subsetting needed).

```{r single-bench, fig.height=4, fig.width=7}
plot_bench_heatmap(
  bench_dfr   = sim_cols$data,
  pheno       = "BNI",
  bench       = "Bench1",
  bench_label = "Bench 1: raw observed BNI",
  show_labels = FALSE
)
```

---

### 5. Adding genotype labels

Set `show_labels = TRUE` to print the genotype name inside each tile. Reduce
`label_size` for large benches where tiles are narrow.

```{r labels, fig.height=4, fig.width=7}
sim_small <- simulate_field_trial(
  n_bench      = 1,
  n_rows       = 5,
  n_cols       = 8,
  spatial_type = 3,
  pheno_name   = "BNI",
  save_csv     = FALSE,
  seed         = 7
)

plot_bench_heatmap(
  bench_dfr   = sim_small$data,
  pheno       = "BNI",
  bench_label = "Bench 1: with genotype labels",
  show_labels = TRUE,
  label_size  = 3
)
```

---

### 6. Shared vs independent colour scales

When `shared_scale = TRUE` (default), the fill colour maps to the global
phenotype range across all benches, making values directly comparable between
panels. When `shared_scale = FALSE`, each bench uses its own range, which can
help reveal within-bench spatial structure when between-bench means differ.

```{r scale-comparison, fig.height=5}
library(patchwork)

global_lims <- range(sim_cols$data$BNI, na.rm = TRUE)

p_shared <- plot_bench_heatmap(
  bench_dfr     = sim_cols$data,
  pheno         = "BNI",
  bench         = "Bench1",
  bench_label   = "Shared scale",
  shared_limits = global_lims
)

p_indep <- plot_bench_heatmap(
  bench_dfr     = sim_cols$data,
  pheno         = "BNI",
  bench         = "Bench1",
  bench_label   = "Independent scale",
  shared_limits = NULL
)

p_shared + p_indep
```

---

### 7. Edge effects and hotspot patterns (types 6 and 7)

Types 6 and 7 are more realistic patterns for glasshouse trials. Type 6 combines a border-proximity effect with a diagonal asymmetry (top-left elevated, bottom-right depressed). Type 7 places a compact Gaussian bell at an off-centre position mimicking a localised disturbance.

```{r sim-edge-hotspot, fig.height=4}
library(patchwork)

sim_edge <- simulate_field_trial(
  spatial_type      = 6,
  spatial_intensity = 30,
  pheno_name        = "BNI",
  save_csv          = FALSE,
  seed              = 42
)

sim_hotspot <- simulate_field_trial(
  spatial_type      = 7,
  spatial_intensity = 30,
  pheno_name        = "BNI",
  save_csv          = FALSE,
  seed              = 42
)

# Plot the true spatial component for Bench1 of each type
lims_edge <- range(sim_edge$data$True_Spatial)
lims_hot  <- range(sim_hotspot$data$True_Spatial)

p_edge <- plot_bench_heatmap(sim_edge$data, pheno = "True_Spatial",
                              bench = "Bench1",
                              bench_label = "Type 6: edge effects",
                              shared_limits = lims_edge)

p_hot <- plot_bench_heatmap(sim_hotspot$data, pheno = "True_Spatial",
                             bench = "Bench1",
                             bench_label = "Type 7: localised hotspot",
                             shared_limits = lims_hot)

p_edge + p_hot
```

The edge-effect pattern is highest at all four borders and lowest in the interior, with a diagonal asymmetry (top-left corner is more elevated than bottom-right). The hotspot is a compact circular region in the upper-right quadrant; outside it the surface is slightly negative (because the pattern is centred to mean zero).

---

### 8. Per-bench spatial intensity

Use `spatial_intensity_per_bench` to assign a different spatial severity to each bench. This is the typical validation design used in the spatial correction report: four benches with intensities 15, 25, 40, 60, ranging from moderate (intensity < genotype SD) to strong (intensity > genotype SD).

```{r sim-per-bench, fig.height=9, fig.width=14}
sim_varying <- simulate_field_trial(
  n_bench                     = 4,
  spatial_type                = 6,
  spatial_intensity           = 30,          # fallback (unused when per_bench supplied)
  spatial_type_per_bench      = rep(6, 4),
  spatial_intensity_per_bench = c(15, 25, 40, 60),
  pheno_name                  = "BNI",
  save_csv                    = FALSE,
  seed                        = 42
)

lims <- range(sim_varying$data$True_Spatial)

panels <- lapply(paste0("Bench", 1:4), function(bn) {
  intensity <- c(15, 25, 40, 60)[as.integer(sub("Bench", "", bn))]
  plot_bench_heatmap(sim_varying$data, pheno = "True_Spatial",
                     bench = bn,
                     bench_label = paste0(bn, " (intensity = ", intensity, ")"),
                     shared_limits = lims)
})

wrap_plots(panels, ncol = 2) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "True spatial pattern: type 6 (edge effects)",
                  subtitle = "Shared colour scale: intensities 15, 25, 40, 60")
```

With a shared colour scale, the 4-fold increase in spatial severity from Bench1 to Bench4 is immediately visible.

---

### 9. Inspecting the true genotype effects

`$true_effects` contains the latent genotype values used to generate the data.
After running `SpatialCorrectionFlexible.R`, the Pearson correlation and RMSE between these true effects and the extracted BLUEs (or BLUPs) quantify how well the spatial correction has recovered the signal.

```{r true-effects, fig.height=3.5, fig.width=5}
head(sim_cols$true_effects)

ggplot(sim_cols$true_effects, aes(x = True_BNI)) +
  geom_histogram(binwidth = 3, fill = "#2c7bb6", colour = "white") +
  labs(title = "Distribution of true genotype effects",
       x = "True BNI", y = "Count") +
  theme_bw()
```

To visualise how true genotype values are arranged spatially across all
benches, merge `$true_effects` with the plot-level data (which carries the
Row, Col, and Bench positions) and pass the result to
`plot_all_bench_heatmaps()`. Because genotypes are randomised independently
per bench, the spatial pattern of true effects will differ between benches,
confirming that the arrangement is random rather than structured.

```{r true-effects-heatmap, fig.height=9, fig.width=14}
all_true <- merge(
  sim_cols$data[, c("Bench", "Genotype", "Row", "Col")],
  sim_cols$true_effects,
  by = "Genotype"
)

p_true <- plot_all_bench_heatmaps(
  dfr          = all_true,
  pheno        = "True_BNI",
  bench_col    = "Bench",
  shared_scale = TRUE,
  ncol_panels  = 2,
  panel_width  = 7,
  panel_height = 4,
  output_png   = "data/heatmap_true_effects.png"
)
p_true
```

---

## Saving outputs

### Output files from this document

All files produced by knitting this document are written to `data/` relative
to the document location (i.e. `docs/data/`):

| File | Contents |
|---|---|
| `BNI_simulation.csv` | Plot-level data for the column-gradient example (section 2) |
| `BNI_true_effects.csv` | Corresponding latent genotype effects |
| `heatmap_type1_none.png` | Composite heatmap: no spatial effect |
| `heatmap_type3_cols.png` | Composite heatmap: column gradient, default layout |
| `heatmap_type3_cols_2col.png` | Composite heatmap: column gradient, 2 panels per row |
| `heatmap_type3_cols_1col.png` | Composite heatmap: column gradient, 1 panel per row |
| `heatmap_true_effects.png` | Composite heatmap: true genotype effects across all benches |

**Note:** Examples using `save_csv = FALSE` (sections 7–8) do not write CSV files.

### Running the script directly

Running `simulate_spatial_data.R` directly with `Rscript` produces the
following files in `data/` at the project root:

| File | Contents |
|---|---|
| `BNI_simulation.csv` | Plot-level data: Bench, Genotype, Row, Col, phenotype value |
| `BNI_true_effects.csv` | Latent genotype effects used to generate the data |
| `heatmaps_raw_labelled.png` | Composite heatmap with genotype names, shared colour scale |

### Saving a single-bench plot

`plot_all_bench_heatmaps()` saves the composite PNG automatically via `ggsave`.
To save a single-bench plot returned by `plot_bench_heatmap()`:

```{r save, eval=FALSE}
p <- plot_bench_heatmap(
  bench_dfr   = sim_cols$data,
  pheno       = "BNI",
  bench       = "Bench1",
  bench_label = "Bench 1"
)

ggsave("data/bench1_heatmap.png", p, width = 9, height = 5, dpi = 150)
```

---

## Session info

```{r session}
sessionInfo()
```
