---
title: "Spatial Correction of Field Trial Data Using GAMM"
subtitle: "Usage guide for `SpatialCorrectionGAMM.R`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5,
  dpi     = 150
)
```

## Overview

`SpatialCorrectionGAMM.R` provides a complete pipeline for removing
environmental spatial gradients from unreplicated or partially replicated
glasshouse and field trial data. It uses penalised regression splines
(via **mgcv**) combined with genotype and positional random effects
(via **nlme**) fitted under REML, producing spatially corrected plot-level
phenotypes and genotype-level Best Linear Unbiased Predictions (BLUPs).

**Experimental design this pipeline is designed for:**

- Multi-bench or multi-environment unreplicated trial: each genotype appears
  **once per bench**, randomly assigned to a plot position
- Bench layout: a regular row × column grid
- Observed phenotype = true genotype value + spatial gradient + random noise

**Outputs:**

| File | Contents |
|---|---|
| `GAMM_plots.pdf` | 6-panel diagnostic page per trait × bench |
| `GAMM_fitted_values.csv` | Plot-level: observed, fitted, residual, and corrected values |
| `GAMM_BLUPs.csv` | Genotype-level BLUPs per bench, plus cross-bench mean rows |
| `Outliers_GAMM_report.txt` | Outlier log with positions and residual values |

---

## Statistical model

### Data-generating process

For bench $b$, each plot $(i, j)$ carrying genotype $k$ contributes an
observed phenotype:

$$y_{ijk} = \mu + g_k + f(r_i,\, c_j) + \rho_i + \kappa_j + \varepsilon_{ijk}$$

where:

| Term | Meaning |
|---|---|
| $\mu$ | Grand mean (fixed intercept) |
| $g_k \sim N(0,\,\sigma_g^2)$ | Genotype random effect |
| $f(r_i, c_j)$ | Fixed spatial surface smooth — captures large-scale environmental gradient |
| $\rho_i \sim N(0,\,\sigma_\rho^2)$ | Row random effect — residual row-level variation not captured by the smooth |
| $\kappa_j \sim N(0,\,\sigma_\kappa^2)$ | Column random effect |
| $\varepsilon_{ijk} \sim N(0,\,\sigma^2)$ | Residual plot-level error |

All random terms are independent. REML estimation via `gamm()` simultaneously
estimates the spatial smooth and the variance components
$\sigma_g^2,\,\sigma_\rho^2,\,\sigma_\kappa^2,\,\sigma^2$.

### The spatial smooth $f(r, c)$

The spatial surface is represented as a **tensor product P-spline**:

$$f(r, c) = \sum_{k_1=1}^{K_r} \sum_{k_2=1}^{K_c}
            \theta_{k_1 k_2}\; B_{k_1}(r) \otimes B_{k_2}(c)$$

where $B_{k_1}(r)$ and $B_{k_2}(c)$ are B-spline basis functions.
Smoothness is enforced by discrete second-difference penalties on
the rows and columns of the coefficient matrix $\Theta$.
The penalty strength is selected automatically by REML.

The number of basis functions ($K_r \times K_c$) is set by `auto_k_dims()`:

$$K_r = \max(3,\; \lfloor n_\text{rows}/2 \rfloor), \quad
  K_c = \max(3,\; \lfloor n_\text{cols}/2 \rfloor)$$

This is conservative — enough knots to represent realistic gradients while
avoiding overfitting. For complex or non-monotone surfaces you can
supply `k_dims` manually.

**Why tensor products rather than isotropic smooths?**
Row and column spacing are typically different, and gradients often differ in
magnitude along rows vs columns (e.g. a strong temperature gradient along the
bench length but little variation across its width). Tensor product splines
apply separate penalties in each direction, so they are more appropriate than
isotropic `s(row, col, bs="tp")` for field-trial layouts.

### Relationship to SpATS / SAP

The model is closely related to the SpATS approach of
Rodriguez-Alvarez et al. (2018) and the P-spline spatial models reviewed
by Piepho et al. (2015). The key differences here are:

- `gamm()` leverages the **nlme** mixed-model engine for REML, giving genuine
  shrinkage BLUPs from `ranef()` rather than approximate predictions
- Row and column random effects provide an additional layer of local correction
  on top of the smooth spatial surface

### Spatial correction and BLUPs

**Corrected plot phenotype.** The mean-centred spatial trend and positional
random effects are subtracted from the observed value:

$$y^{*}_{ijk} = y_{ijk} - \bigl[\hat{f}(r_i,c_j) - \overline{\hat{f}}\bigr]
               - \hat{\rho}_i - \hat{\kappa}_j$$

This leaves the grand mean, genotype BLUP, and residual error intact,
producing corrected values on the original phenotype scale.

**Genotype BLUPs.** The adjusted genotype mean is:

$$\hat{\mu}_k = \hat{\mu} + \hat{g}_k$$

where $\hat{g}_k$ is the BLUP from `ranef(model$lme)[[gt_col]]`.

**Reliability** (shrinkage coefficient):

$$h^2_r = \frac{\sigma_g^2}{\sigma_g^2 + \sigma^2 / n_\text{rep}}$$

A reliability near 1 means the BLUP is close to the raw mean; near 0 means
heavy shrinkage toward the grand mean due to low genotypic signal.

---

## Function reference

### Helper functions

#### `auto_k_dims(mydfr, rn, cn)`

Automatically selects the number of tensor-product spline basis functions
for a given bench.

| Argument | Description |
|---|---|
| `mydfr` | Data frame for a single bench |
| `rn` | Name of the row column |
| `cn` | Name of the column column |

Returns `c(k_row, k_col)`: integer vector used as the `k` argument to `te()`.

---

#### `fit_gamm_spatial(mydfr, pheno, gt, rn, cn, k_dims)`

Fits the GAMM model for one bench × one trait.
Attaches the prepared data frame as an attribute for downstream use.

| Argument | Description |
|---|---|
| `mydfr` | Single-bench data frame |
| `pheno` | Name of the response column |
| `gt` | Name of the genotype column |
| `rn` | Name of the row column |
| `cn` | Name of the column column |
| `k_dims` | Integer vector `c(k_row, k_col)` from `auto_k_dims()` |

Returns a `gamm()` object with components `$gam` and `$lme`.

---

#### `detect_outliers_residuals(model, mydfr, rn, cn, method, k)`

Flags outliers from the model's level-1 (within-group) conditional residuals.
These residuals have the genotype and smooth contributions removed, so only
genuine measurement errors or data entry mistakes should be flagged.

| Argument | Default | Description |
|---|---|---|
| `model` | — | `gamm()` object |
| `mydfr` | — | Data frame used to fit the model |
| `rn` | — | Row column name |
| `cn` | — | Column column name |
| `method` | `"IQR"` | `"IQR"`: flag beyond $Q_1 - k \cdot \text{IQR}$ or $Q_3 + k \cdot \text{IQR}$; `"SD"`: flag beyond mean $\pm\, k \cdot \text{SD}$ |
| `k` | `3` | Threshold multiplier (3 is conservative) |

Returns a logical vector (length = rows in `mydfr`); `TRUE` = outlier.

---

#### `extract_blups_gamm(model, mydfr, gt, n_rep)`

Extracts genotype BLUPs, standard error, and reliability from a fitted model.

| Argument | Default | Description |
|---|---|---|
| `model` | — | `gamm()` object |
| `mydfr` | — | Data frame (must have `gt` column as a factor) |
| `gt` | — | Genotype column name |
| `n_rep` | `1` | Replicates per genotype per bench (1 for unreplicated designs) |

Returns a data frame: `Genotype`, `predicted`, `StdErr`, `reliability`.

---

#### `compute_variogram(resids, rows, cols, n_bins, max_pairs)`

Computes an empirical variogram of the model residuals.
Used diagnostically: a flat variogram after correction confirms no
remaining spatial autocorrelation in the residuals.

| Argument | Default | Description |
|---|---|---|
| `resids` | — | Numeric residual vector |
| `rows` | — | Row positions |
| `cols` | — | Column positions |
| `n_bins` | `15` | Number of distance bins |
| `max_pairs` | `5000` | Maximum random pairs evaluated (for speed) |

Returns a data frame: `dist_bin_mid`, `semivariance`, `n_pairs`.

---

#### `plot_spatial_diagnostics(model, mydfr, blups, pheno, bench_label, rn, cn, gt)`

Builds a six-panel **patchwork** diagnostic page for one trait × one bench.

| Panel | Content | What to look for |
|---|---|---|
| 1 | Observed phenotype heatmap | Raw spatial pattern in the data |
| 2 | Fitted values heatmap | Smooth representing genotype + spatial fit |
| 3 | Residuals heatmap | Should be unstructured (no spatial pattern) |
| 4 | Fitted spatial surface on a fine grid | Shape and scale of the estimated gradient |
| 5 | BLUPs ranked dot-plot with ±1 SE bars | Spread of genotype effects; SE uniformity |
| 6 | Empirical variogram of residuals | Should flatten quickly; sustained rise indicates residual spatial signal |

Returns a patchwork ggplot object.

---

### Main pipeline function

#### `run_spatial_gamm(mydfr, pheno, gt, rn, cn, bench_label, pmain, outlier_method, outlier_k)`

Full pipeline for one bench × one trait:

1. Auto-select `k_dims`
2. Fit an initial GAMM
3. Flag outliers on conditional residuals; replace with `NA`
4. Re-fit GAMM on cleaned data
5. Extract BLUPs and per-plot corrected values
6. Print a 6-panel diagnostic page to the active graphics device

| Argument | Default | Description |
|---|---|---|
| `mydfr` | — | Single-bench data frame |
| `pheno` | — | Response column name |
| `gt` | — | Genotype column name |
| `rn` | — | Row column name |
| `cn` | — | Column column name |
| `bench_label` | — | String identifier for this bench (used in titles and output) |
| `pmain` | — | Plot title prefix |
| `outlier_method` | `"IQR"` | Outlier detection method (`"IQR"` or `"SD"`) |
| `outlier_k` | `3` | Multiplier for outlier threshold |

Returns `list(dfrF, dfrP)`:

- `dfrF`: plot-level data frame with `*_Observed`, `*_Fitted`, `*_Residual`,
  `*_Corrected` columns
- `dfrP`: BLUP data frame with `*_StdErr` and `*_Reliability` columns

---

#### `plot_smooth_heatmap(model, mydfr, rn, cn, bench_label, pheno_label)`

Standalone plot of the fitted spatial surface on the **actual plot grid**
(one tile per plot, not interpolated). Uses a diverging blue–white–red
colour scale centred at zero:

- **Blue tiles** — plots suppressed by the environment; their observed value
  was lower than their true genotype effect; the correction adds to the
  observed value
- **Red tiles** — plots elevated by the environment; the correction subtracts
  from the observed value

| Argument | Default | Description |
|---|---|---|
| `model` | — | `gamm()` object (per-bench or pooled) |
| `mydfr` | — | Data frame used to fit the model |
| `rn` | — | Row column name |
| `cn` | — | Column column name |
| `bench_label` | `""` | Label shown in the plot title |
| `pheno_label` | `"Spatial\ntrend"` | Fill legend label |

Returns a `ggplot` object.

Note: this function is called automatically inside `run_spatial_gamm()` and
`run_spatial_gamm_pooled()` after the 6-panel diagnostics.

---

#### `fit_gamm_pooled(dfr, pheno, gt, rn, cn, bench_col, k_dims)`

Fits one GAMM for all benches simultaneously with a **shared** spatial smooth.
`Bench` is included as a fixed effect to absorb between-bench mean differences.
Row and column random effects are bench-specific (named `Bench_rN` and
`Bench_cN`) to prevent row 1 in bench A from being treated as the same group
as row 1 in bench B.

Use this when you have good reason to believe the spatial gradient has the
same shape on every bench (e.g. identical bench construction, same orientation
relative to heating/lighting). When patterns differ, prefer `fit_gamm_spatial()`
called separately per bench.

---

#### `extract_blups_gamm_pooled(model, dfr, gt, bench_col)`

Extracts genotype BLUPs from a pooled model. The grand mean is computed as
the intercept plus the **average of all bench fixed effects**, so that the
reported BLUP is not anchored to any single reference bench.

---

#### `run_spatial_gamm_pooled(dfr, pheno, gt, rn, cn, bench_col, outlier_method, outlier_k)`

Full pipeline for the shared-smooth model across all benches simultaneously.
Produces three diagnostic plots: the shared smooth heatmap, a BLUP dotplot,
and a variogram of the pooled residuals. Returns `list(dfrF, dfrP)` where
`dfrP$Bench = "Pooled"`.

---

### Controlling which benches are processed

Both the script and the Rmd tutorial support partial bench processing via
`bench_filter`. This is useful for quick testing or when some benches have
data quality issues.

```r
bench_filter <- NULL          # (default) process all benches
bench_filter <- c("Bench1")   # process only Bench1
bench_filter <- c("Bench1", "Bench3")  # process a subset
```

When `bench_col = NULL` (no bench structure in the data), `bench_filter` has
no effect and all data are processed as a single field.

---

### Choosing between separate and shared smoothers

| Setting | When to use |
|---|---|
| `separate_smoothers = TRUE` (default) | Different benches may have different spatial gradients. Each bench gets its own `te(Row, Col)` surface estimated independently. No assumption about pattern similarity across benches. |
| `separate_smoothers = FALSE` | You have reason to believe all benches share the same spatial gradient (same equipment, same orientation). Pooling borrows strength across benches and produces a single cross-bench BLUP directly. |

---

### Using the script directly

Edit the **CONFIG** section near the bottom of `SpatialCorrectionGAMM.R`:

```r
fn          <- "path/to/your_data.csv"   # input file
fnpdf       <- "GAMM_plots.pdf"
report_file <- "Outliers_GAMM_report.txt"
fnFitted    <- "GAMM_fitted_values.csv"
fnPredicted <- "GAMM_BLUPs.csv"

gt_col    <- "Genotype"    # column name for genotypes
rn_col    <- "Row"         # column name for row positions
cn_col    <- "Column"      # column name for column positions
bench_col <- "Bench"       # set to NULL for a single field with no bench column
trFc      <- 5             # column index of the FIRST trait (traits run from here to the end)

outlier_method <- "IQR"    # "IQR" or "SD"
outlier_k      <- 3        # multiplier (3 is conservative)

bench_filter       <- NULL   # NULL = all benches; e.g. c("Bench1","Bench2") to process a subset
separate_smoothers <- TRUE   # TRUE = separate smooth per bench (default); FALSE = shared smooth
```

Then run: `Rscript scripts/SpatialCorrectionGAMM.R`

The CONFIG section is skipped when the script is sourced from R or knitted.

---

## Tutorial — Example 1: Uniform column gradient

### Scenario

Four glasshouse benches, each containing 150 plots (10 rows × 15 columns).
A half-sine temperature gradient runs along the length of each bench
(column direction), with identical magnitude and shape on all benches.
The spatial effect peaks at ~25 BNI units, comparable in scale to the
between-genotype SD (12 units) — a realistic but challenging correction
problem.

We will:

1. Simulate the data and visualise the raw spatial patterns
2. Run the spatial GAMM per bench
3. Examine the diagnostic plots
4. Extract BLUPs and compare them to the known true genotype effects

### Setup

```{r source-scripts}
source("../scripts/simulate_spatial_data.R")
source("../scripts/SpatialCorrectionGAMM.R")

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```

### Simulate data

```{r ex1-simulate}
sim1 <- simulate_field_trial(
  n_bench           = 4,
  n_rows            = 10,
  n_cols            = 15,
  mean_pheno        = 55,
  sd_geno           = 12,
  spatial_type      = 3,          # column gradient
  spatial_intensity = 25,
  spatial_scale     = 1,
  sd_error          = 4,
  pheno_name        = "BNI",
  save_csv          = TRUE,
  output_dir        = "data",
  seed              = 42
)

cat("Plots:     ", nrow(sim1$data), "\n")
cat("Genotypes: ", length(unique(sim1$data$Genotype)), "\n")
cat("Benches:   ", length(unique(sim1$data$Bench)), "\n")
head(sim1$data)
```

### Raw observed heatmaps

The column gradient is clearly visible as a systematic left-to-right colour
ramp in each bench. Because genotypes are randomly assigned within each bench,
the underlying genotype differences create local noise on top of this trend.

```{r ex1-raw-heatmap, fig.width=14, fig.height=9}
plot_all_bench_heatmaps(
  dfr          = sim1$data,
  pheno        = "BNI",
  bench_col    = "Bench",
  shared_scale = TRUE,
  ncol_panels  = 2,
  panel_width  = 7,
  panel_height = 4,
  output_png   = "data/ex1_raw_heatmap.png"
)
```

### True genotype effect distribution

```{r ex1-true-effects, fig.width=6, fig.height=4}
ggplot(sim1$true_effects, aes(x = True_BNI)) +
  geom_histogram(binwidth = 3, fill = "#2c7bb6", colour = "white") +
  geom_vline(xintercept = mean(sim1$true_effects$True_BNI),
             colour = "firebrick", linetype = "dashed") +
  labs(title = "Distribution of true genotype effects",
       x = "True BNI", y = "Count") +
  theme_bw()
```

### Run spatial GAMM — one bench at a time

Each call to `run_spatial_gamm()` fits the full pipeline for one bench and
automatically prints a 6-panel diagnostic page. Progress messages are shown
below each run.

```{r ex1-gamm, fig.width=14, fig.height=8}
benches_ex1    <- unique(sim1$data$Bench)
results_ex1    <- list()
fitted_ex1     <- NULL

for (b in benches_ex1) {

  bench_data <- sim1$data[sim1$data$Bench == b, ]

  res <- run_spatial_gamm(
    mydfr          = bench_data,
    pheno          = "BNI",
    gt             = "Genotype",
    rn             = "Row",
    cn             = "Col",
    bench_label    = b,
    pmain          = paste("Ex1", b),
    outlier_method = "IQR",
    outlier_k      = 3
  )

  if (!is.null(res)) {
    results_ex1[[b]] <- res
    fitted_ex1 <- if (is.null(fitted_ex1)) res$dfrF else rbind(fitted_ex1, res$dfrF)
  }
}
```

### Inspect the spatial smooth heatmap

After the 6-panel diagnostic, `run_spatial_gamm()` automatically prints a
discrete smooth heatmap on the actual plot grid. You can also call
`plot_smooth_heatmap()` directly on any fitted model:

```{r ex1-smooth-demo, eval=FALSE}
# Example: plot the spatial smooth for one bench directly
bench_data_b1 <- sim1$data[sim1$data$Bench == "Bench1", ]
k_dims_b1     <- auto_k_dims(bench_data_b1, "Row", "Col")
model_b1      <- fit_gamm_spatial(bench_data_b1, "BNI", "Genotype", "Row", "Col", k_dims_b1)
plot_smooth_heatmap(model_b1, bench_data_b1, "Row", "Col", bench_label = "Bench1")
```

The smooth heatmap from the GAMM run appears inline above (printed by
`run_spatial_gamm()` automatically). The column gradient should be visible
as a shift from blue (left columns, suppressed) to red (right columns, elevated).

---

### Interpreting the 6-panel diagnostic

For a well-fitted model you should see:

- **Panel 1 (Observed):** the raw gradient visible in the heatmap
- **Panel 2 (Fitted):** a smooth approximation of Panel 1 — not a perfect
  copy, since genotype differences are absorbed into the random effects
- **Panel 3 (Residuals):** spatially unstructured noise — no clear gradient
  or banding
- **Panel 4 (Spatial surface):** the smooth $\hat{f}(r, c)$ on a fine grid.
  For a column gradient this should show monotonically increasing colour from
  left to right, with little variation across rows
- **Panel 5 (BLUPs ranked):** a spread of adjusted genotype means. Narrow
  standard error bars indicate stable estimation
- **Panel 6 (Variogram):** semivariance should level off quickly with distance.
  A steadily rising variogram would indicate unmodelled spatial structure in
  the residuals

### Examine BLUPs

Collect per-bench BLUPs and compute a cross-bench mean for each genotype.

```{r ex1-blups}
blup_list_ex1 <- lapply(results_ex1, function(x) x$dfrP)
blups_ex1     <- do.call(rbind, blup_list_ex1)

# Cross-bench mean BLUP (unweighted)
mean_blups_ex1 <- blups_ex1 %>%
  group_by(Genotype) %>%
  summarise(BNI_BLUP_mean = mean(BNI, na.rm = TRUE), .groups = "drop")

cat("Number of genotypes with BLUPs:", nrow(mean_blups_ex1), "\n")
summary(mean_blups_ex1$BNI_BLUP_mean)
```

```{r ex1-blup-plot, fig.width=8, fig.height=4}
mean_blups_ex1_ord <- mean_blups_ex1[order(mean_blups_ex1$BNI_BLUP_mean), ]
mean_blups_ex1_ord$rank <- seq_len(nrow(mean_blups_ex1_ord))

ggplot(mean_blups_ex1_ord, aes(x = rank, y = BNI_BLUP_mean)) +
  geom_point(size = 1.5, colour = "#2c7bb6") +
  labs(title = "Cross-bench mean BLUPs — Example 1",
       x = "Genotype rank", y = "Mean BLUP (BNI)") +
  theme_bw()
```

### Compare BLUPs to true effects

The `BNI_SpatialTrend` column in the fitted data records the exact spatial
adjustment applied to each plot. Visualising this confirms the model captured
the column gradient:

```{r ex1-spatial-trend, fig.width=14, fig.height=5}
library(viridis)

# Plot SpatialTrend for all benches in a shared colour scale
trend_lim <- max(abs(fitted_ex1$BNI_SpatialTrend), na.rm = TRUE)

ggplot(fitted_ex1, aes(x = col, y = row, fill = BNI_SpatialTrend)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_gradientn(
    colours = c("#2166ac", "white", "#d6604d"),
    limits  = c(-trend_lim, trend_lim),
    name    = "Spatial\ntrend"
  ) +
  scale_y_reverse() +
  facet_wrap(~ Bench, nrow = 1) +
  labs(title = "Fitted spatial trend applied per bench",
       subtitle = "Blue = plots adjusted upward; red = adjusted downward",
       x = "Col", y = "Row") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

We can quantify correction performance by correlating the estimated BLUPs
against the known true genotype values that were used to generate the data.
We also compute the raw-mean correlation as a baseline (what you would get
without any correction).

```{r ex1-performance}
# Raw means (per-genotype average of observed values across benches)
raw_means_ex1 <- sim1$data %>%
  group_by(Genotype) %>%
  summarise(raw_mean = mean(BNI, na.rm = TRUE), .groups = "drop")

# Corrected means (per-genotype average of spatially corrected values)
corrected_means_ex1 <- fitted_ex1 %>%
  group_by(Genotype) %>%
  summarise(corrected_mean = mean(BNI_Corrected, na.rm = TRUE), .groups = "drop")

# Combine with true effects
perf_ex1 <- sim1$true_effects %>%
  inner_join(raw_means_ex1,      by = "Genotype") %>%
  inner_join(corrected_means_ex1, by = "Genotype") %>%
  inner_join(mean_blups_ex1,     by = "Genotype")

cor_raw       <- cor(perf_ex1$raw_mean,       perf_ex1$True_BNI, use = "complete.obs")
cor_corrected <- cor(perf_ex1$corrected_mean, perf_ex1$True_BNI, use = "complete.obs")
cor_blup      <- cor(perf_ex1$BNI_BLUP_mean,  perf_ex1$True_BNI, use = "complete.obs")

cat(sprintf("Pearson r — raw means vs true:       %.3f\n", cor_raw))
cat(sprintf("Pearson r — corrected means vs true: %.3f\n", cor_corrected))
cat(sprintf("Pearson r — BLUPs vs true:           %.3f\n", cor_blup))
```

The improvement in correlation from raw means to BLUPs quantifies how
much of the spatial confounding has been removed. With a peak spatial effect
of ~25 BNI units against a genotype SD of 12, this is a genuinely challenging
correction.

```{r ex1-scatter, fig.width=12, fig.height=5}
p_raw <- ggplot(perf_ex1, aes(x = True_BNI, y = raw_mean)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#d7301f") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_raw)) +
  labs(title = "Raw means vs true effects",
       x = "True BNI", y = "Raw mean across benches") +
  theme_bw()

p_blup <- ggplot(perf_ex1, aes(x = True_BNI, y = BNI_BLUP_mean)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_blup)) +
  labs(title = "BLUPs vs true effects",
       x = "True BNI", y = "Mean BLUP across benches") +
  theme_bw()

p_raw + p_blup
```

A well-fitting correction should move the scatter plot from the left panel
(where points fan out due to spatial confounding) to the right panel (where
points cluster tightly around the diagonal).

### Contrast: per-bench vs pooled for a uniform pattern

For a uniform spatial pattern the per-bench and pooled models should perform
equivalently — they are fitting the same surface, just parameterised
differently. We verify this here. Example 2 will show what happens when the
patterns are **not** uniform.

```{r ex1-pooled-fit, fig.width=14, fig.height=6}
res_pooled_ex1 <- run_spatial_gamm_pooled(
  dfr            = sim1$data,
  pheno          = "BNI",
  gt             = "Genotype",
  rn             = "Row",
  cn             = "Col",
  bench_col      = "Bench",
  outlier_method = "IQR",
  outlier_k      = 3
)
```

```{r ex1-pooled-perf, fig.width=12, fig.height=5}
blups_pooled_ex1 <- res_pooled_ex1$dfrP

perf_pooled_ex1 <- sim1$true_effects %>%
  inner_join(blups_pooled_ex1[, c("Genotype", "BNI")], by = "Genotype") %>%
  rename(BNI_BLUP_pooled = BNI)

cor_pooled_ex1 <- cor(perf_pooled_ex1$BNI_BLUP_pooled,
                       perf_pooled_ex1$True_BNI, use = "complete.obs")

cat(sprintf("BLUP r — per-bench (separate smoothers): %.3f\n", cor_blup))
cat(sprintf("BLUP r — pooled   (shared smoother):    %.3f\n", cor_pooled_ex1))
cat("\nFor a uniform gradient the two approaches are near-equivalent.\n")

p_sep1 <- ggplot(perf_ex1, aes(x = True_BNI, y = BNI_BLUP_mean)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_blup)) +
  labs(title = "Per-bench (separate smoothers)", x = "True BNI", y = "BLUP mean") +
  theme_bw()

p_pool1 <- ggplot(perf_pooled_ex1, aes(x = True_BNI, y = BNI_BLUP_pooled)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#7fbc41") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_pooled_ex1)) +
  labs(title = "Pooled (shared smoother)", x = "True BNI", y = "Pooled BLUP") +
  theme_bw()

p_sep1 + p_pool1 +
  plot_annotation(title = "Example 1 (uniform gradient): per-bench ≈ pooled")
```

---

## Phenotype outputs for downstream analysis (GWAS)

### BLUP, BLUE and de-regressed BLUP — what is the difference?

All three values are in the `dfrP` output table. They serve different downstream purposes:

| Value | Column | Calculation | Use |
|---|---|---|---|
| **BLUP** | `BNI` | grand mean + shrunken genotype RE | Ranking, selection, genetic parameters |
| **BLUE** | `BNI_BLUE` | mean spatially corrected observation | GWAS with kinship matrix; unbiased, no shrinkage |
| **de-regressed BLUP** | `BNI_deBLUP` | grand mean + ranef / reliability | GWAS tools expecting de-regressed values + weights |

The **reliability** column (`BNI_Reliability`) is the shrinkage factor:
$$\text{reliability} = \frac{\sigma_g^2}{\sigma_g^2 + \sigma_e^2 / n_{\text{rep}}}$$

For a GWAS with a kinship matrix (EMMAX, GCTA, BGLR):
$$\hat{g}_k^{\text{BLUE}} = \bar{y}_k^{\text{corrected}} \quad \text{(mean spatially corrected value across benches)}$$

For de-regressed BLUPs (Garrick et al. 2009, used in animal breeding GWAS):
$$\hat{g}_k^{\text{deBLUP}} = \mu + \frac{\hat{u}_k}{\text{reliability}_k}$$

where $\hat{u}_k$ is the raw (unscaled) random effect deviation. De-regressed BLUPs are weighted in GWAS by $w_k = \text{reliability}_k / (1 - \text{reliability}_k)$.

### When to use which

```{r gwas-table, echo=FALSE}
gwas_df <- data.frame(
  `GWAS workflow` = c(
    "Single-step (kinship model on raw data)",
    "Two-step: kinship model on pre-adjusted phenotype",
    "Two-step: no kinship (simple regression per marker)",
    "Selection / ranking",
    "Heritability / genetic parameter estimation"
  ),
  `Recommended output` = c(
    "BNI_Corrected from dfrF (plot-level)",
    "BNI_BLUE (cross-bench mean, no shrinkage)",
    "BNI_BLUE (unbiased, but population structure uncorrected)",
    "BNI (BLUP — shrunken, best ranking accuracy)",
    "BNI (BLUP reliabilities inform heritability)"
  ),
  `Do NOT use` = c(
    "—",
    "BNI (BLUP — shrinkage attenuates marker effects)",
    "BNI (BLUP — double-shrinkage)",
    "BNI_BLUE (noisier than BLUP for ranking)",
    "—"
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(gwas_df)
```

### Performance comparison: BLUP vs BLUE vs de-regressed BLUP

```{r gwas-comparison, fig.width=13, fig.height=5}
# Collect cross-bench means for all three output types from Example 1
blup_list_ex1 <- lapply(results_ex1, function(x) x$dfrP)
blups_all_ex1 <- do.call(rbind, blup_list_ex1)

gwas_means <- blups_all_ex1 %>%
  group_by(Genotype) %>%
  summarise(
    BLUP   = mean(BNI,          na.rm = TRUE),
    BLUE   = mean(BNI_BLUE,     na.rm = TRUE),
    deBLUP = mean(BNI_deBLUP,  na.rm = TRUE),
    .groups = "drop"
  )

gwas_perf <- sim1$true_effects %>%
  inner_join(gwas_means, by = "Genotype")

r_blup   <- round(cor(gwas_perf$BLUP,   gwas_perf$True_BNI, use = "complete.obs"), 3)
r_blue   <- round(cor(gwas_perf$BLUE,   gwas_perf$True_BNI, use = "complete.obs"), 3)
r_deblup <- round(cor(gwas_perf$deBLUP, gwas_perf$True_BNI, use = "complete.obs"), 3)

cat(sprintf("Pearson r vs true effect — BLUP:   %.3f\n", r_blup))
cat(sprintf("Pearson r vs true effect — BLUE:   %.3f\n", r_blue))
cat(sprintf("Pearson r vs true effect — deBLUP: %.3f\n", r_deblup))

# Show the three options side by side
gwas_long <- gwas_perf %>%
  tidyr::pivot_longer(c(BLUP, BLUE, deBLUP),
                      names_to = "Type", values_to = "Value") %>%
  mutate(Type = factor(Type, levels = c("BLUP", "BLUE", "deBLUP")))

r_ann_gwas <- data.frame(
  Type  = factor(c("BLUP", "BLUE", "deBLUP"), levels = c("BLUP", "BLUE", "deBLUP")),
  label = paste0("r = ", c(r_blup, r_blue, r_deblup))
)

ggplot(gwas_long, aes(x = True_BNI, y = Value)) +
  geom_point(alpha = 0.4, size = 1.5, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "#d7301f", linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50", linewidth = 0.5) +
  geom_text(data = r_ann_gwas, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.5, size = 3.5, inherit.aes = FALSE) +
  facet_wrap(~ Type, nrow = 1) +
  labs(
    title    = "BLUP vs BLUE vs de-regressed BLUP — Example 1 (uniform gradient)",
    subtitle = paste0("All three recover the true effect well (r ≥ ", min(r_blup, r_blue, r_deblup),
                      "). BLUPs are shrunk toward the mean; BLUEs and deBLUPs lie on the y=x line."),
    x = "True BNI", y = "Estimated value"
  ) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold", size = 10))
```

Notice the slope of the regression line (red) relative to the identity (dashed):

- **BLUP**: slope < 1 — shrinkage compresses values toward the mean. The BLUP is a better *predictor* (lower MSE) but a biased estimate of the true effect magnitude.
- **BLUE / deBLUP**: slope ≈ 1 — unbiased. These lie closer to the identity line and are the correct input for GWAS where the kinship matrix handles population structure.

For GWAS, use the **cross-bench mean BLUE** (`BNI_BLUE` averaged across benches, computed above as `gwas_means$BLUE`). If your GWAS software requires reliability weights (e.g. `wt_GWAS = reliability / (1 - reliability)`), use `BNI_deBLUP` with those weights.

---

## Tutorial — Example 2: Heterogeneous spatial patterns

### Scenario

In practice, different benches often have different spatial structures —
a bench near a heating vent may show a localised warm patch, while a bench
in the open aisle shows a smooth end-to-end gradient.

In this example we simulate a four-bench trial where:

- **Bench_Grad1 and Bench_Grad2** have a column gradient (same as Example 1)
- **Bench_Patch1 and Bench_Patch2** have a central Gaussian hot-spot

Because `simulate_field_trial()` is deterministic (given the same seed,
grid dimensions, and genotype parameters), calling it twice with the same
`seed`, `n_bench`, `n_rows`, and `n_cols` but different `spatial_type`
produces:

- **Identical true genotype effects** (both calls draw `geno_true` from
  the same RNG state)
- **Identical within-bench plot assignments** and noise vectors
- **Different spatial surfaces** only

This means Example 2 uses the exact same 150 genotypes with the same
underlying values as Example 1 — only the environmental context differs
in benches 3 and 4.

### The modelling question

Example 1 showed that per-bench and pooled models give near-identical results
when all benches share the same spatial structure. In that situation the pooled
model is slightly more parsimonious.

This example tests the **heterogeneous case**. We will fit both models and
compare:

- **Per-bench model** — separate `te(Row, Col)` for each bench, estimated
  independently. Adapts freely to any pattern.
- **Pooled model** — one shared `te(Row, Col)` across all benches. Only valid
  if all benches have the same spatial structure.

The per-bench model should excel here; the pooled model should underperform
because it must fit a single compromise surface over four genuinely different
spatial patterns.

### Simulate mixed-pattern data

```{r ex2-simulate}
# Benches 1-2: column gradient
sim_grad <- simulate_field_trial(
  n_bench = 2, n_rows = 10, n_cols = 15,
  mean_pheno = 55, sd_geno = 12,
  spatial_type = 3, spatial_intensity = 25,
  sd_error = 4, pheno_name = "BNI",
  save_csv = FALSE, seed = 42
)

# Benches 3-4: central Gaussian patch (same seed → same genotype effects)
sim_patch <- simulate_field_trial(
  n_bench = 2, n_rows = 10, n_cols = 15,
  mean_pheno = 55, sd_geno = 12,
  spatial_type = 5, spatial_intensity = 25,
  sd_error = 4, pheno_name = "BNI",
  save_csv = FALSE, seed = 42
)

# Rename bench identifiers so they are unique after combining
sim_grad$data$Bench  <- gsub("Bench", "Bench_Grad",  sim_grad$data$Bench)
sim_patch$data$Bench <- gsub("Bench", "Bench_Patch", sim_patch$data$Bench)

sim2_data    <- rbind(sim_grad$data, sim_patch$data)
true_efx2    <- sim_grad$true_effects   # identical to sim_patch$true_effects

cat("Total plots:", nrow(sim2_data), "\n")
cat("Bench IDs: ", paste(unique(sim2_data$Bench), collapse = ", "), "\n")
```

### Visualise raw data — different patterns clearly visible

```{r ex2-raw-heatmap, fig.width=14, fig.height=9}
plot_all_bench_heatmaps(
  dfr          = sim2_data,
  pheno        = "BNI",
  bench_col    = "Bench",
  shared_scale = TRUE,
  ncol_panels  = 2,
  panel_width  = 7,
  panel_height = 4,
  output_png   = "data/ex2_raw_heatmap.png"
)
```

The top two panels show the left-to-right gradient (Bench_Grad1 and
Bench_Grad2), while the bottom two panels show a bright central area fading
toward the edges (Bench_Patch1 and Bench_Patch2). The colour scale is shared,
so intensity differences reflect real phenotype differences.

### Run the spatial GAMM — same pipeline, heterogeneous data

The call to `run_spatial_gamm()` is identical to Example 1. No special
configuration is needed for the mixed-pattern benches.

```{r ex2-gamm, fig.width=14, fig.height=8}
benches_ex2 <- unique(sim2_data$Bench)
results_ex2 <- list()
fitted_ex2  <- NULL

for (b in benches_ex2) {

  bench_data <- sim2_data[sim2_data$Bench == b, ]

  res <- run_spatial_gamm(
    mydfr          = bench_data,
    pheno          = "BNI",
    gt             = "Genotype",
    rn             = "Row",
    cn             = "Col",
    bench_label    = b,
    pmain          = paste("Ex2", b),
    outlier_method = "IQR",
    outlier_k      = 3
  )

  if (!is.null(res)) {
    results_ex2[[b]] <- res
    fitted_ex2 <- if (is.null(fitted_ex2)) res$dfrF else rbind(fitted_ex2, res$dfrF)
  }
}
```

### What the diagnostic plots reveal

Examine **Panel 4 (Spatial surface)** across the four benches:

- **Bench_Grad1 and Bench_Grad2:** the surface should show a monotone ramp
  from left (low) to right (high), confirming the model captured the
  column gradient
- **Bench_Patch1 and Bench_Patch2:** the surface should show a central
  elevated region fading toward the edges — the model adapts its shape to
  match the Gaussian hot-spot without any user input

**Panel 3 (Residuals)** should look spatially unstructured in all four cases,
confirming that the spatial trend has been successfully removed regardless of
its shape.

**Panel 6 (Variogram)** should flatten quickly for all benches. Any sustained
upward slope would suggest residual spatial autocorrelation — a sign that
the smooth complexity (`k_dims`) may need to be increased.

### Compare BLUPs to true effects

```{r ex2-performance}
blup_list_ex2 <- lapply(results_ex2, function(x) x$dfrP)
blups_ex2     <- do.call(rbind, blup_list_ex2)

mean_blups_ex2 <- blups_ex2 %>%
  group_by(Genotype) %>%
  summarise(BNI_BLUP_mean = mean(BNI, na.rm = TRUE), .groups = "drop")

raw_means_ex2 <- sim2_data %>%
  group_by(Genotype) %>%
  summarise(raw_mean = mean(BNI, na.rm = TRUE), .groups = "drop")

corrected_means_ex2 <- fitted_ex2 %>%
  group_by(Genotype) %>%
  summarise(corrected_mean = mean(BNI_Corrected, na.rm = TRUE), .groups = "drop")

perf_ex2 <- true_efx2 %>%
  inner_join(raw_means_ex2,       by = "Genotype") %>%
  inner_join(corrected_means_ex2, by = "Genotype") %>%
  inner_join(mean_blups_ex2,      by = "Genotype")

cor_raw_ex2  <- cor(perf_ex2$raw_mean,       perf_ex2$True_BNI, use = "complete.obs")
cor_corr_ex2 <- cor(perf_ex2$corrected_mean, perf_ex2$True_BNI, use = "complete.obs")
cor_blup_ex2 <- cor(perf_ex2$BNI_BLUP_mean,  perf_ex2$True_BNI, use = "complete.obs")

cat(sprintf("Pearson r — raw means vs true:       %.3f\n", cor_raw_ex2))
cat(sprintf("Pearson r — corrected means vs true: %.3f\n", cor_corr_ex2))
cat(sprintf("Pearson r — BLUPs vs true:           %.3f\n", cor_blup_ex2))
```

```{r ex2-scatter, fig.width=12, fig.height=5}
p_raw2 <- ggplot(perf_ex2, aes(x = True_BNI, y = raw_mean)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#d7301f") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_raw_ex2)) +
  labs(title = "Raw means vs true effects",
       x = "True BNI", y = "Raw mean across benches") +
  theme_bw()

p_blup2 <- ggplot(perf_ex2, aes(x = True_BNI, y = BNI_BLUP_mean)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_blup_ex2)) +
  labs(title = "BLUPs vs true effects",
       x = "True BNI", y = "Mean BLUP across benches") +
  theme_bw()

p_raw2 + p_blup2
```

### Spatial trend heatmaps — different patterns per bench

```{r ex2-trend-heatmap, fig.width=14, fig.height=5}
trend_lim2 <- max(abs(fitted_ex2$BNI_SpatialTrend), na.rm = TRUE)

ggplot(fitted_ex2, aes(x = col, y = row, fill = BNI_SpatialTrend)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_gradientn(
    colours = c("#2166ac", "white", "#d6604d"),
    limits  = c(-trend_lim2, trend_lim2),
    name    = "Spatial\ntrend"
  ) +
  scale_y_reverse() +
  facet_wrap(~ Bench, nrow = 1) +
  labs(title = "Fitted spatial trend — per-bench model",
       subtitle = "Left benches: column gradient captured  |  Right benches: central patch captured",
       x = "Col", y = "Row") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

The heatmaps confirm that the per-bench model has correctly identified
different spatial structures: a monotone left-to-right ramp for the gradient
benches, and a central hot-spot for the patch benches.

### Contrast: shared smooth vs per-bench smooth on mixed-pattern data

When patterns genuinely differ, fitting a **shared smooth** across all benches
produces a compromise surface that fits none of them well. The code below fits
the pooled model on the same mixed-pattern data and compares BLUP recovery.

```{r ex2-pooled-contrast, fig.width=6, fig.height=4}
# Pool all mixed-pattern data and fit one shared smooth
keep2 <- c("BNI", "Genotype", "Row", "Col", "Bench")
sim2_pooled_input <- sim2_data[, keep2]

res_pooled_ex2 <- run_spatial_gamm_pooled(
  dfr            = sim2_pooled_input,
  pheno          = "BNI",
  gt             = "Genotype",
  rn             = "Row",
  cn             = "Col",
  bench_col      = "Bench",
  outlier_method = "IQR",
  outlier_k      = 3
)
```

```{r ex2-pooled-perf}
blups_pooled_ex2 <- res_pooled_ex2$dfrP  # Bench = "Pooled"

perf_pooled_ex2 <- true_efx2 %>%
  inner_join(blups_pooled_ex2[, c("Genotype", "BNI")], by = "Genotype") %>%
  rename(BNI_BLUP_pooled = BNI)

cor_pooled_ex2 <- cor(perf_pooled_ex2$BNI_BLUP_pooled,
                       perf_pooled_ex2$True_BNI,
                       use = "complete.obs")

cat(sprintf("BLUP r — per-bench model (separate smoothers): %.3f\n", cor_blup_ex2))
cat(sprintf("BLUP r — pooled model (shared smoother):       %.3f\n", cor_pooled_ex2))
cat("\nWhen patterns differ across benches, the per-bench model\n")
cat("typically recovers genotype effects more accurately.\n")
```

```{r ex2-pooled-scatter, fig.width=12, fig.height=5}
p_sep <- ggplot(perf_ex2, aes(x = True_BNI, y = BNI_BLUP_mean)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_blup_ex2)) +
  labs(title = "Separate smoothers (per bench)",
       x = "True BNI", y = "BLUP mean") +
  theme_bw()

p_pool <- ggplot(perf_pooled_ex2, aes(x = True_BNI, y = BNI_BLUP_pooled)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#d7301f") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_pooled_ex2)) +
  labs(title = "Shared smoother (pooled)",
       x = "True BNI", y = "Pooled BLUP") +
  theme_bw()

p_sep + p_pool +
  plot_annotation(title = "Mixed-pattern data: separate vs shared spatial smooth")
```

### Summary across all examples

```{r summary-table}
summary_df <- data.frame(
  Scenario = c(
    "Ex1: Uniform gradient — per-bench",
    "Ex1: Uniform gradient — pooled",
    "Ex2: Mixed patterns   — per-bench",
    "Ex2: Mixed patterns   — pooled"
  ),
  Pearson_r_raw  = round(c(cor_raw,     cor_raw,        cor_raw_ex2,   cor_raw_ex2),   3),
  Pearson_r_BLUP = round(c(cor_blup,    cor_pooled_ex1, cor_blup_ex2,  cor_pooled_ex2), 3),
  stringsAsFactors = FALSE
)
print(summary_df, row.names = FALSE)
```

**Key takeaways:**

- For a **uniform spatial pattern** (Example 1), per-bench and pooled models
  give near-identical BLUP recovery — both are valid; the pooled model is
  more parsimonious.
- For **heterogeneous patterns** (Example 2), the per-bench model adapts to
  each bench independently and substantially outperforms the pooled model,
  which is forced to fit a compromise surface over incompatible structures.
- The per-bench approach is therefore the **safe default**: it costs nothing
  extra when patterns are uniform, but is essential when they differ.

---

## Should we loop over benches?

### The problem with per-bench looping

The default pipeline fits a **separate model for each bench** and then
averages the per-bench BLUPs:

$$\hat{g}_k^{\text{per-bench}} = \frac{1}{n_b} \sum_{l=1}^{n_b}
  \bigl(\hat{\mu}_l + \hat{u}_{k|l}\bigr)$$

This is a **simple average**, not a mixed-model BLUP. It has three
statistical weaknesses:

1. **Variance components estimated independently per bench.** The genotype
   variance $\sigma_g^2$ is re-estimated from scratch on each bench's
   ~150 observations. Estimates are noisy, and are not shared across benches
   even though the same genotypes are observed on all benches.

2. **No genotype × bench interaction (GEI) component.** A single model
   across benches can decompose the variance into $\sigma_g^2$ (stable
   genotype main effect) and $\sigma_{gb}^2$ (genotype–bench interaction).
   The per-bench loop conflates these, producing BLUPs that are mixtures
   of both.

3. **No formal cross-bench shrinkage.** If a genotype's observation is
   missing or an outlier on one bench, its cross-bench mean BLUP is pulled
   toward that bench's data equally. In a proper joint model, the missing
   bench contributes nothing and the BLUP borrows only from the benches
   that have data.

### The joint multi-bench model

The statistically appropriate model for a multi-bench unreplicated trial
pools all benches in a single REML fit:

$$y_{ijkl} = \mu + \beta_l + f_l(r_i, c_j) + g_k + \rho_{il} + \kappa_{jl}
             + \varepsilon_{ijkl}$$

| Term | Meaning |
|---|---|
| $\beta_l$ | Bench fixed effect (absorbs between-bench mean differences) |
| $f_l(r, c)$ | Bench-specific spatial surface — separate `te()` per bench via `by = Bench` |
| $g_k \sim N(0, \sigma_g^2)$ | Genotype random effect estimated jointly across all benches |
| $\rho_{il}, \kappa_{jl}$ | Bench-specific row / column random effects |

The critical improvement is that **$\sigma_g^2$ is estimated from all bench
data simultaneously**, and each genotype BLUP is shrunk toward the grand
mean using the pooled estimate. The `te(Row, Col, by = Bench)` term allows
completely different spatial surfaces per bench — identical flexibility to
separate models — but within a single likelihood.

**When bench as random is preferred.**
If benches are drawn from a population (e.g. multiple glasshouse experiments
in different years), bench can be treated as a random effect:
$\beta_l \sim N(0, \sigma_b^2)$. This makes BLUPs portable to
unobserved benches but requires removing the parametric bench term, which
complicates smooth identification. For a fixed, known set of benches
(e.g. a single experiment), the fixed-bench formulation above is
appropriate and avoids this complication.

**A note on GEI in unreplicated designs.**
With exactly 1 replicate per genotype per bench, the
genotype × bench interaction $\sigma_{gb}^2$ is confounded with
residual error $\sigma^2$. Including a `s(Genotype, Bench, bs="re")`
interaction term will cause the residual to collapse to zero.
For unreplicated designs, it is better to omit GEI and acknowledge that
$\hat{g}_k$ captures the average genotype effect across benches —
variation in rank across benches is absorbed into the residual.

### Implementation: joint multi-bench GAM via `mgcv::gam()`

We use `mgcv::gam()` with `bs = "re"` smooths for random effects (the
"GAMM via GAM" approach, Wood 2017 §6.6). This keeps all computations in
the `mgcv` package without calling `nlme` or `lme4`.

```{r define-mb-functions}
#' Fit joint multi-bench GAM with bench-specific spatial smooths
#' @param dfr       full data frame (all benches)
#' @param pheno     response column name
#' @param gt        genotype column name
#' @param rn        row column name
#' @param cn        column column name
#' @param bench_col bench column name
#' @param k_dims    integer vector c(k_row, k_col)
#' @return gam() object
fit_gam_multibench <- function(dfr, pheno, gt, rn, cn, bench_col, k_dims) {
  dfr$bench_rowf   <- as.factor(paste0(dfr[[bench_col]], "_r", dfr[[rn]]))
  dfr$bench_colf   <- as.factor(paste0(dfr[[bench_col]], "_c", dfr[[cn]]))
  dfr[[bench_col]] <- as.factor(dfr[[bench_col]])
  dfr[[gt]]        <- as.factor(dfr[[gt]])

  # te(Row, Col, by=Bench): separate P-spline surface per bench
  # + Bench (fixed): required for smooth identifiability with factor by-variable
  # s(Genotype, bs="re"): genotype random intercept — estimated jointly
  # s(bench_rowf/colf, bs="re"): bench-specific row/col random intercepts
  formula_str <- paste0(
    pheno, " ~ ", bench_col,
    " + te(", rn, ", ", cn, ", by=", bench_col,
    ", bs=c('ps','ps'), k=c(", k_dims[1], ",", k_dims[2], "))",
    " + s(", gt, ", bs='re')",
    " + s(bench_rowf, bs='re') + s(bench_colf, bs='re')"
  )
  gam(as.formula(formula_str), data = dfr, method = "REML")
}

#' Extract genotype BLUPs from a joint multi-bench GAM
#' @param model     gam() object from fit_gam_multibench()
#' @param dfr       data frame used for fitting
#' @param gt        genotype column name
#' @param bench_col bench column name
#' @return data.frame: Genotype, predicted
extract_blups_gam_mb <- function(model, dfr, gt, bench_col) {
  # predict(type="terms") gives the contribution of each smooth/parametric
  # term per observation.  The genotype smooth s(gt) is a random intercept:
  # its value is the same for every observation belonging to the same genotype.
  terms_mat <- predict(model, type = "terms")
  gt_col_nm <- paste0("s(", gt, ")")
  gt_re     <- terms_mat[, gt_col_nm]

  # Average over plots per genotype (all values identical for a given genotype)
  geno_re <- tapply(gt_re, dfr[[gt]], mean)

  # Grand mean: parametric intercept + mean bench fixed effect
  fe        <- coef(model)[!grepl("^s\\(", names(coef(model)))]
  intercept <- as.numeric(fe["(Intercept)"])
  bench_fe  <- fe[grepl(paste0("^", bench_col), names(fe))]
  grand_mean <- intercept + mean(c(0, as.numeric(bench_fe)))

  data.frame(
    Genotype  = names(geno_re),
    predicted = grand_mean + as.numeric(geno_re),
    stringsAsFactors = FALSE
  )
}
```

```{r fit-mb}
k_dims_mb <- auto_k_dims(sim1$data[sim1$data$Bench == "Bench1", ], "Row", "Col")

model_mb <- fit_gam_multibench(
  dfr       = sim1$data,
  pheno     = "BNI",
  gt        = "Genotype",
  rn        = "Row",
  cn        = "Col",
  bench_col = "Bench",
  k_dims    = k_dims_mb
)

# Brief model summary — effective degrees of freedom for each term
cat("k_dims:", k_dims_mb, "\n")
cat("Terms and EDF:\n")
print(round(summary(model_mb)$edf, 2))
cat("Total EDF:", round(sum(model_mb$edf), 1), "\n")
```

```{r mb-blups}
blups_mb <- extract_blups_gam_mb(model_mb, sim1$data, "Genotype", "Bench")

# Compare directly to true effects
comp_mb <- merge(blups_mb, sim1$true_effects, by = "Genotype")
cor_mb  <- cor(comp_mb$predicted, comp_mb$True_BNI)

cat(sprintf("Joint multi-bench GAM — BLUP r vs true: %.3f\n", cor_mb))
cat(sprintf("Per-bench GAMM (loop)  — BLUP r vs true: %.3f\n", cor_blup))
```

```{r mb-scatter, fig.width=12, fig.height=5}
# Merge per-bench and joint BLUPs for side-by-side scatter
comp_both <- merge(mean_blups_ex1, blups_mb, by = "Genotype") %>%
  inner_join(sim1$true_effects, by = "Genotype")

p_loop <- ggplot(comp_both, aes(x = True_BNI, y = BNI_BLUP_mean)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_blup)) +
  labs(title = "Per-bench loop (averaged BLUPs)",
       x = "True BNI", y = "BLUP") +
  theme_bw()

p_joint <- ggplot(comp_both, aes(x = True_BNI, y = predicted)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#1a9641") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = sprintf("r = %.3f", cor_mb)) +
  labs(title = "Joint multi-bench GAM",
       x = "True BNI", y = "BLUP") +
  theme_bw()

p_loop + p_joint +
  plot_annotation(title = "Per-bench loop vs joint model — Example 1 (column gradient)")
```

The joint model is expected to produce marginally better or equal recovery.
For a well-controlled glasshouse trial the two approaches typically agree
closely, since σ²_g is reasonably stable across benches. In field trials
or multi-year experiments with large GEI, the joint model will show a
clearer advantage.

---

## Comparison with SpATS

### Background

**SpATS** (Spatial Analysis of Trials with Splines; Rodriguez-Alvarez et al.
2018) fits a similar P-spline spatial correction model to the GAMM, using
an efficient REML algorithm based on the SAP (Smoothing and P-splines)
framework. The main structural differences are:

| Feature | GAMM (`mgcv`) | SpATS |
|---|---|---|
| Spatial surface | `te(row, col, bs=c("ps","ps"))` | `SAP(col, row)` or `PSANOVA(col, row)` |
| Random effects | `nlme::lme()` / `bs="re"` | Internal REML (Henderson equations) |
| Multi-bench | Separate models or joint via `by=` | Typically per-bench; no native joint model |
| BLUP output | `ranef()` / `predict(type="terms")` | `predict(model, which=gt)` |
| Spatial plot | Manual via `plot_smooth_heatmap()` | `plot(model)` — built-in |

SpATS is specifically designed for this application and tends to be faster
per bench. The GAMM approach is more flexible (arbitrary random effects,
multi-bench pooling, richer diagnostics).

### SpATS per-bench fitting

```{r spats-check}
has_spats <- requireNamespace("SpATS", quietly = TRUE)
if (!has_spats) message("SpATS not installed — install.packages('SpATS') to run this section.")
```

```{r define-spats-fn}
#' Fit SpATS model for one bench, returning genotype BLUPs and spatial trend
#' @param bench_data  single-bench data frame
#' @param pheno       response column name
#' @param gt          genotype column name
#' @param rn          row column name
#' @param cn          column column name
#' @param k_row       approximate row complexity (nseg_row = max(4, k_row-1))
#' @param k_col       approximate col complexity (nseg_col = max(4, k_col-1))
#' @return list with $blups (Genotype, predicted, se) and
#'         $plot_data (Row, Col, Fitted, SpatialTrend)
run_spats_bench <- function(bench_data, pheno, gt, rn, cn, k_row, k_col) {
  if (!requireNamespace("SpATS", quietly = TRUE))
    stop("SpATS package required")

  d        <- bench_data
  d$rn_var <- d[[rn]]
  d$cn_var <- d[[cn]]
  d$rnf    <- as.factor(d$rn_var)
  d$cnf    <- as.factor(d$cn_var)
  d[[gt]]  <- as.factor(d[[gt]])

  n_seg_col <- max(4L, as.integer(k_col) - 1L)
  n_seg_row <- max(4L, as.integer(k_row) - 1L)

  invisible(capture.output(
    spatD <- SpATS::SpATS(
      response           = pheno,
      spatial            = ~ SpATS::SAP(cn_var, rn_var,
                                         nseg = c(n_seg_col, n_seg_row)),
      genotype           = gt,
      genotype.as.random = TRUE,
      random             = ~ rnf + cnf,
      data               = d,
      control            = list(tolerance = 1e-04, monitoring = 0)
    )
  ))

  geno_pred <- SpATS::predict.SpATS(spatD, which = gt)

  # Spatial trend = fitted values minus the per-plot genotype BLUP, centred
  geno_blup_vec <- geno_pred$predicted.values[
    match(as.character(d[[gt]]), as.character(geno_pred[[gt]]))]
  spatial_raw   <- spatD$fitted - geno_blup_vec
  spatial_trend <- spatial_raw - mean(spatial_raw, na.rm = TRUE)

  list(
    blups = data.frame(
      Genotype  = as.character(geno_pred[[gt]]),
      predicted = geno_pred$predicted.values,
      se        = geno_pred$standard.errors,
      stringsAsFactors = FALSE
    ),
    plot_data = data.frame(
      Row          = d[[rn]],
      Col          = d[[cn]],
      Fitted       = spatD$fitted,
      SpatialTrend = spatial_trend,
      stringsAsFactors = FALSE
    )
  )
}
```

```{r fit-spats, eval=has_spats}
# Fit SpATS on Example 1 (uniform gradient) — all 4 benches
k_s1 <- auto_k_dims(sim1$data[sim1$data$Bench == "Bench1", ], "Row", "Col")

spats_ex1 <- lapply(unique(sim1$data$Bench), function(b) {
  bd  <- sim1$data[sim1$data$Bench == b, ]
  res <- tryCatch(
    run_spats_bench(bd, "BNI", "Genotype", "Row", "Col", k_s1[1], k_s1[2]),
    error = function(e) { message("SpATS Ex1 bench ", b, ": ", e$message); NULL }
  )
  if (is.null(res)) return(NULL)
  res$blups$Bench     <- b
  res$plot_data$Bench <- b
  res
})
spats_ex1 <- Filter(Negate(is.null), spats_ex1)

spats_blups_ex1  <- do.call(rbind, lapply(spats_ex1, `[[`, "blups"))
spats_plots_ex1  <- do.call(rbind, lapply(spats_ex1, `[[`, "plot_data"))
spats_means_ex1  <- spats_blups_ex1 %>%
  group_by(Genotype) %>%
  summarise(BNI_SpATS = mean(predicted, na.rm = TRUE), .groups = "drop")

# Fit SpATS on Example 2 (mixed patterns) — all 4 benches
k_s2 <- auto_k_dims(sim2_data[sim2_data$Bench == "Bench_Grad1", ], "Row", "Col")

spats_ex2 <- lapply(unique(sim2_data$Bench), function(b) {
  bd  <- sim2_data[sim2_data$Bench == b, ]
  res <- tryCatch(
    run_spats_bench(bd, "BNI", "Genotype", "Row", "Col", k_s2[1], k_s2[2]),
    error = function(e) { message("SpATS Ex2 bench ", b, ": ", e$message); NULL }
  )
  if (is.null(res)) return(NULL)
  res$blups$Bench     <- b
  res$plot_data$Bench <- b
  res
})
spats_ex2 <- Filter(Negate(is.null), spats_ex2)

spats_blups_ex2 <- do.call(rbind, lapply(spats_ex2, `[[`, "blups"))
spats_plots_ex2 <- do.call(rbind, lapply(spats_ex2, `[[`, "plot_data"))
spats_means_ex2 <- spats_blups_ex2 %>%
  group_by(Genotype) %>%
  summarise(BNI_SpATS = mean(predicted, na.rm = TRUE), .groups = "drop")

cat("SpATS: Ex1 BLUPs for", nrow(spats_means_ex1), "genotypes,",
    "Ex2 BLUPs for", nrow(spats_means_ex2), "genotypes\n")
```

---

## GAMM vs SpATS comparison

### Spatial surface comparison — Example 1, one bench

The most direct way to compare the two methods is to visualise what each
fitted as the spatial trend for the same bench. Both use P-spline surfaces;
minor differences reflect algorithm choices (GAMM via `nlme` REML vs SpATS
SAP).

```{r spats-surface-compare, fig.width=14, fig.height=5, eval=has_spats}
# Extract GAMM spatial trend for Bench1
gamm_trend_b1 <- fitted_ex1[fitted_ex1$Bench == "Bench1",
                              c("row", "col", "BNI_SpatialTrend")]
names(gamm_trend_b1) <- c("Row", "Col", "SpatialTrend")
gamm_trend_b1$Method <- "GAMM te(Row,Col)"

# SpATS spatial trend for Bench1
spats_trend_b1 <- spats_plots_ex1[spats_plots_ex1$Bench == "Bench1",
                                    c("Row", "Col", "SpatialTrend")]
spats_trend_b1$Method <- "SpATS SAP(Col,Row)"

surf_both <- rbind(gamm_trend_b1, spats_trend_b1)
lim_surf  <- max(abs(surf_both$SpatialTrend), na.rm = TRUE)

ggplot(surf_both, aes(x = Col, y = Row, fill = SpatialTrend)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_gradientn(
    colours = c("#2166ac", "white", "#d6604d"),
    limits  = c(-lim_surf, lim_surf),
    name    = "Spatial\ntrend"
  ) +
  scale_y_reverse() +
  facet_wrap(~ Method) +
  labs(title = "Estimated spatial trend — Bench 1 (column gradient)",
       subtitle = "Both methods should produce near-identical left-to-right ramps") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

### BLUP agreement — do the two methods rank genotypes the same way?

```{r spats-blup-agree, fig.width=11, fig.height=5, eval=has_spats}
# Combine per-bench GAMM and SpATS BLUPs for Example 1
agree_ex1 <- mean_blups_ex1 %>%
  rename(GAMM = BNI_BLUP_mean) %>%
  inner_join(rename(spats_means_ex1, SpATS = BNI_SpATS), by = "Genotype")

r_agree_ex1 <- round(cor(agree_ex1$GAMM, agree_ex1$SpATS), 3)

# Same for Example 2
agree_ex2 <- mean_blups_ex2 %>%
  rename(GAMM = BNI_BLUP_mean) %>%
  inner_join(rename(spats_means_ex2, SpATS = BNI_SpATS), by = "Genotype")

r_agree_ex2 <- round(cor(agree_ex2$GAMM, agree_ex2$SpATS), 3)

p_ag1 <- ggplot(agree_ex1, aes(x = GAMM, y = SpATS)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#2c7bb6") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "#d7301f", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = paste0("r = ", r_agree_ex1)) +
  labs(title = "Ex1: Uniform gradient",
       x = "GAMM BLUP", y = "SpATS BLUP") +
  theme_bw()

p_ag2 <- ggplot(agree_ex2, aes(x = GAMM, y = SpATS)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#7fbc41") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "#d7301f", linewidth = 0.7) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.5,
           label = paste0("r = ", r_agree_ex2)) +
  labs(title = "Ex2: Mixed patterns",
       x = "GAMM BLUP", y = "SpATS BLUP") +
  theme_bw()

(p_ag1 + p_ag2) +
  plot_annotation(
    title    = "GAMM vs SpATS per-bench BLUPs — genotype agreement",
    subtitle = "Points on the dashed identity line = perfect agreement; red line = fitted regression"
  )
```

### Full performance comparison across both examples

```{r build-comparison-df, eval=has_spats}
cor_spats_ex1 <- cor(
  inner_join(sim1$true_effects, spats_means_ex1, by = "Genotype")$BNI_SpATS,
  inner_join(sim1$true_effects, spats_means_ex1, by = "Genotype")$True_BNI
)
cor_spats_ex2 <- cor(
  inner_join(true_efx2, spats_means_ex2, by = "Genotype")$BNI_SpATS,
  inner_join(true_efx2, spats_means_ex2, by = "Genotype")$True_BNI
)

perf_full <- data.frame(
  Scenario = rep(c("Ex1: Uniform gradient", "Ex2: Mixed patterns"), each = 4),
  Model = rep(c("Raw means", "GAMM per-bench", "GAMM pooled", "SpATS per-bench"), 2),
  Pearson_r = round(c(
    cor_raw,        cor_blup,     cor_pooled_ex1, cor_spats_ex1,
    cor_raw_ex2,    cor_blup_ex2, cor_pooled_ex2, cor_spats_ex2
  ), 3)
)
print(perf_full, row.names = FALSE)
```

```{r comparison-bar, fig.width=9, fig.height=5, eval=has_spats}
perf_full$Model <- factor(perf_full$Model,
  levels = c("Raw means", "GAMM pooled", "SpATS per-bench", "GAMM per-bench"))

ggplot(perf_full, aes(x = Model, y = Pearson_r, fill = Model)) +
  geom_col(width = 0.6, colour = "white") +
  geom_text(aes(label = sprintf("%.3f", Pearson_r)),
            vjust = -0.4, size = 3.2) +
  scale_fill_manual(values = c(
    "Raw means"       = "#d7301f",
    "GAMM pooled"     = "#7fbc41",
    "SpATS per-bench" = "#f1a340",
    "GAMM per-bench"  = "#2c7bb6"
  ), guide = "none") +
  scale_y_continuous(limits = c(0, 1.08), expand = c(0, 0)) +
  facet_wrap(~ Scenario) +
  labs(title = "BLUP recovery: Pearson r vs true genotype effects",
       x = NULL, y = "Pearson r") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid.major.x = element_blank())
```

```{r comparison-scatter, fig.width=14, fig.height=5, eval=has_spats}
# Build long-format data for faceted scatter — Example 1
comp_ex1_long <- sim1$true_effects %>%
  inner_join(mean_blups_ex1[, c("Genotype", "BNI_BLUP_mean")] %>%
               rename(`GAMM per-bench` = BNI_BLUP_mean), by = "Genotype") %>%
  inner_join(perf_pooled_ex1[, c("Genotype", "BNI_BLUP_pooled")] %>%
               rename(`GAMM pooled` = BNI_BLUP_pooled), by = "Genotype") %>%
  inner_join(spats_means_ex1[, c("Genotype", "BNI_SpATS")] %>%
               rename(`SpATS per-bench` = BNI_SpATS), by = "Genotype") %>%
  tidyr::pivot_longer(-c(Genotype, True_BNI), names_to = "Model", values_to = "BLUP") %>%
  mutate(Scenario = "Ex1: Uniform gradient")

comp_ex2_long <- true_efx2 %>%
  inner_join(mean_blups_ex2[, c("Genotype", "BNI_BLUP_mean")] %>%
               rename(`GAMM per-bench` = BNI_BLUP_mean), by = "Genotype") %>%
  inner_join(perf_pooled_ex2[, c("Genotype", "BNI_BLUP_pooled")] %>%
               rename(`GAMM pooled` = BNI_BLUP_pooled), by = "Genotype") %>%
  inner_join(spats_means_ex2[, c("Genotype", "BNI_SpATS")] %>%
               rename(`SpATS per-bench` = BNI_SpATS), by = "Genotype") %>%
  tidyr::pivot_longer(-c(Genotype, True_BNI), names_to = "Model", values_to = "BLUP") %>%
  mutate(Scenario = "Ex2: Mixed patterns")

comp_long <- bind_rows(comp_ex1_long, comp_ex2_long)

r_ann <- comp_long %>%
  group_by(Scenario, Model) %>%
  summarise(r = round(cor(BLUP, True_BNI, use = "complete.obs"), 3), .groups = "drop") %>%
  mutate(label = paste0("r=", r))

ggplot(comp_long, aes(x = True_BNI, y = BLUP)) +
  geom_point(alpha = 0.3, size = 1, colour = "#2c7bb6") +
  geom_smooth(method = "lm", se = FALSE, colour = "#d7301f", linewidth = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50", linewidth = 0.4) +
  geom_text(data = r_ann, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.5, size = 3, inherit.aes = FALSE) +
  facet_grid(Scenario ~ Model) +
  labs(title = "All methods vs true genotype effects",
       subtitle = "Dashed = identity line (perfect recovery)",
       x = "True BNI", y = "Estimated BLUP") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold", size = 8))
```

### Interpretation

- **Spatial surfaces** (GAMM vs SpATS): both identify the same gradient or
  patch structure. Minor tile-level differences reflect different REML
  algorithms (GAMM via `nlme`, SpATS via SAP) rather than a fundamental
  modelling difference.
- **BLUP agreement**: GAMM and SpATS per-bench BLUPs are highly correlated
  (typically r > 0.99 within-bench). Differences in absolute scale reflect
  differences in shrinkage — SpATS uses a slightly different variance
  component estimation scheme.
- **Which to use?** For standard unreplicated glasshouse trials either
  method is appropriate. GAMM (this pipeline) offers richer random-effect
  flexibility and integrates into the pooled/joint multi-bench framework.
  SpATS is faster per bench and has a built-in spatial diagnostic plot.
- **The critical choice** is not GAMM vs SpATS but **per-bench vs pooled**:
  both methods show that the per-bench approach is robust to heterogeneous
  patterns while pooled is not.

---

## Output files from this document

All files are written to `data/` relative to the document location (`docs/data/`).

| File | Contents |
|---|---|
| `BNI_simulation.csv` | Example 1 plot-level data (column gradient, 4 benches) |
| `BNI_true_effects.csv` | True genotype effects for Example 1 |
| `ex1_raw_heatmap.png` | Composite raw heatmap — Example 1 |
| `ex2_raw_heatmap.png` | Composite raw heatmap — Example 2 (mixed patterns) |

**Packages required:**
`mgcv`, `nlme`, `ggplot2`, `dplyr`, `tidyr`, `viridis`, `patchwork`,
`SpATS` (optional — SpATS comparison section skipped if not installed).

---

## References

- Rodriguez-Alvarez, M.X., Boer, M.P., van Eeuwijk, F.A., Eilers, P.H.C.
  (2018). Correcting for spatial heterogeneity in plant breeding experiments
  with P-splines. *Theoretical and Applied Genetics*, 131, 703–716.
- Piepho, H.-P., Möhring, J., Pflugfelder, M., Hermann, W., Williams, E.R.
  (2015). Problems in parameter estimation for power and AR(1) models of
  spatial correlation in designed field experiments.
  *Communications in Biometry and Crop Science*, 10, 3–16.
- Wood, S.N. (2017). *Generalized Additive Models: An Introduction with R*
  (2nd ed.). Chapman and Hall/CRC.
- Rodriguez-Alvarez, M.X., Boer, M.P., van Eeuwijk, F.A., Eilers, P.H.C.
  (2018). SpATS: Spatial Analysis of Trials with Splines. R package.
  CRAN.

---

## Session info

```{r session}
sessionInfo()
```
